PyTracer: Automatically profiling numerical instabilities in Python

### 2021-12-10 paper.tex
Greg review

\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:} #1\color{black}\xspace}
\newcommand{\tristanmod}[2]{\color{orange}\sout{#1} #2\color{black}\xspace}

@@

\title{PyTracer: Automatically profiling numerical instabilities in Python}
\author{Yohan Chatelain$^1$, Gregory Kiar$^2$, Nigel Young$^1$, Tristan Glatard$^1$\\
$^1$Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada\\
$^2$Center for the Developing Brain, Child Mind Institute, New York, NY, USA}
\date{}
\maketitle

\begin{abstract}
Numerical stability is a crucial requirement of reliable scientific computing. However, despite the pervasiveness of Python in data science, analyzing large Python programs remains challenging due to the lack of scalable numerical analysis tools available for this language. We developed \pytracer, a profiler to quantify numerical instability in Python applications. \pytracer automatically instruments Python code to produce numerical traces and visualize them interactively in a Plotly dashboard. We designed \pytracer to work with any numerical noise model including Monte-Carlo Arithmetic, random rounding or random data perturbations. We illustrate \pytracer's capabilities by testing the numerical stability of key functions in the SciPy and Scikit-learn Python libraries. Through these evaluations, we demonstrate \pytracer as a scalable, automatic, and generic framework for numerical evaluations in Python.
% Numerical stability is a crucial requirement of reliable scientific computing. However, despite the pervasiveness of Python in data science, analyzing large Python programs remains challenging due to the lack of scalable numerical analysis tools available for this language. To address this issue, we developed \pytracer, a profiler to quantify numerical instability in Python applications. \pytracer automatically instruments Python code to produce numerical traces and visualize them interactively in a Plotly dashboard.
% \pytracer is designed to work with any type of numerical noise model, including Monte-Carlo Arithmetic, random rounding or random data perturbations. We illustrate \pytracer's capabilities by testing the numerical stability of key functions in the SciPy and Scikit-learn Python libraries. Finally, \pytracer slows down executions by only $3.5 \times$ on average.
% \tristan{Add one sentence to summarize the findings} 
\end{abstract}

@@

The scientific Python ecosystem is a central component of data
analyses owing to its rich offering of data manipulation, array programming,
numerical analysis, and visualization tools. In particular, libraries such
as NumPy~\cite{harris2020array}, SciPy~\cite{virtanen2020scipy}, scikit-learn~\cite{pedregosa2011scikit} or PyTorch~\cite{paszke2019pytorch} are used in hundreds of publications every year, providing a reference set of high-quality open-source core scientific libraries. Researchers have built numerous domain-specific software tools from this ecosystem, leveraging Python's simplicity and expressivity.

Numerical stability is a crucial requirement of reliable scientific data
processing. In unstable analyses, small numerical perturbations introduced by data noise, software and hardware updates, or parallelization lead to substantial deviations in final results and potentially different scientific conclusions, threatening the reliability of computational analyses. To address this issue, numerical stability analyses need to be conducted systematically, however, no practical tool currently exists to conduct such analyses for Python programs.

We present PyTracer, a numerical stability analyzer for Python programs.
PyTracer adopts a dynamic approach that evaluates numerical stability using program execution traces and is applicable to large programs without manual intervention. In contrast, theoretical approaches based on condition numbers or backward error analysis require detailed modeling of the program and its numerical implementation. Similarly, static code analysis techniques hardly scale to large codebases, particularly in dynamically-typed languages such as Python.

PyTracer estimates the variability of floating-point variables by combining program traces obtained with different numerical perturbations, which requires (1) generating numerically-perturbed executions, (2) tracing floating-point computations along program executions, and (3) visualizing stability evaluations. PyTracer addresses these requirements with (1) a ``fuzzy" Python interpreter instrumented with Monte-Carlo arithmetic, (2) a dynamic instrumentation of Python functions, modules, and classes through meta-programming, (3) an interactive Plotly dashboard to visualize stability measures throughout program executions.

@@

Several tools have been developed to assess numerical quality, divided into static and dynamic approaches. Static approaches typically analyze source code while dynamic approaches trace floating-point computations throughout executions. A detailed review of these approaches is presented in~\cite{cherubin2020tools}. 
\pytracer adopts the dynamic approach due to its scalability to large existing code bases. Dynamic approaches typically trace floating-point computations, detect instabilities in traces, and visualize summary statistics about instabilities. This section reviews existing approaches for each of these steps.

@@

\label{sec:soa}
\subsection{Numerical tracing techniques}
Numerous tools exists to trace floating-point computations for C, C++, or FORTRAN programs due to the prevalence of these languages in High-Performance Computing (HPC). 
The main tracing techniques are source-to-source, compiler-based transformations, and dynamic binary instrumentation.

The source-to-source technique requires a rewriting of the application to modify floating-point types. 
CADNA~\cite{jezequel2008cadna} is a library for C, C++, and Fortran implementing the CESTAC~\cite{vignes1993stochastic} stochastic arithmetic. Shaman~\cite{demeure_phd} is a C++ library that uses a first-order error model to propagate numerical error. 
MCAlib~\cite{frechtling2015mcalib} is a C library that implements the Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} using the MPFR~\cite{fousse2007mpfr} library.
Finally, the work in~\cite{tang2016software} proposed a source-to-source framework for executing targeted code in infinite and fixed precision with and without  stochastic arithmetic.

The compiler-based approach instruments floating-point expressions at compile time. 
Verificarlo~\cite{verificarlo} is a compiler that supports different types of arithmetic instrumentations, including MCA. 
The work in~\cite{bao2013fly} modified the GNU Compiler Collection (GCC) to track local floating-point errors across executions. pLiner~\cite{guo2020pliner} is a root-cause analyzer based on the Clang compiler that detects floating-point operations responsible for result variability using a source-to-source transformation at the Abstract Syntax Tree (AST) level to rewrite parts of code with higher precision. 
PFPSanitizer~\cite{chowdhary2020debugging,chowdhary2021parallel} is a compiler that uses parallel shadow execution to detect numerical issues using higher precision.
FLiT~\cite{sawaya2017flit} is a framework to detect variability induced by compilers and their optimizations.
The work in~\cite{wang2012development} proposes a numerical debugger based on GDB~\cite{stallman1988debugging} for Discrete Stochastic Arithmetic (DSA) on FPGA as an Eclipse plugging. Similarly, Cadtrace~\cite{jezequel2008cadna} and Shaman propose a GDB-based tool to use the CADNA and Shaman libraries with GDB, respectively.

@@

Being agnostic to the programming language is a serious advantage of the DBI method compared to source-to-source and compiler-bases methods. However, working at the binary level makes it difficult to access high-level information needed to debug and understand the source-code logic. Conversely, the source-to-source approach provides a fine-grained control on the analyzed code, but it lacks scalability, as rewriting large codes can be a tedious task. Finally, the compiler-based approach takes advantage of the best of both by being automatic and having access to high-level information. However, like source-to-source, the compiler-based method is not suitable for analyzing closed-source libraries.

@@

Three main approaches exist to detect numerical instabilities: stochastic arithmetic, uncertainty or sensitivity analysis~\cite{helton2006survey}, and random seed analysis~\cite{click2011quality}.
%\tristan{add key references for uq/sa if any}
\pytracer focuses on stochastic arithmetic since it does not make any assumption about 
the noise model, in contrast to sensitivity analysis.
%\tristan{due to...}
Stochastic arithmetic leverages randomness to estimate numerical instabilities coming from the use of floating-point representations. The main idea is to treat round-off errors 
% \tristan{only round-off errors or all numerical errors?} 
as random variables and characterize them statistically.

@@

Stochastic arithmetic quantifies numerical error using the number of significant digits $s$ estimated among the sampled values. A common formula to determine this number from MCA samples is the one reported in~\cite{parker1997monte}:
\begin{equation}
s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a variable sampled with MCA.  
Sohier et al.~\cite{sohier2018confidence} recently provided a generalization of this formula to include confidence intervals.

@@

Visualization covers a wide range of numerical simulation domains, from general-purpose numerical visualization with tools
such as VTK~\cite{schroeder2000visualizing} to topological data analysis~\cite{tierny2018topological}, including matrix~\cite{wu2008matrix} or tensor~\cite{kindlmann2006diffusion} visualization.
Despite the large number of existing tools to identify numerically unstable sections of source code, few allow the user to have a global view of the numerical precision of their computations or to compare several executions of the same program under various perturbations.

Veritracer~\cite{chatelain2018veritracer} was the first tool to propose an embedded visualizer of numerical stability over the execution of C, C++, and Fortran codes. Veritracer extends Verificarlo~\cite{verificarlo} with contextual information about traced variables, which allows the user to identify instabilities quickly.
However, Veritracer neither offers a dynamic interface nor includes information about high-level structures such as \texttt{struct} or \texttt{array}.
FPAvizual~\cite{gu2014fpavisual} visualizes the effects of floating-point arithmetic but only for educational purposes.
Anteater~\cite{faust2019anteater} proposes a visualizer for Python codes, but as mentioned previously, it only deals with native Python types, and the user needs to select variables to trace manually. Anteater provides an insightful reflection 
around visual debugging beyond floating-point analysis, which has inspired the design of \pytracer's visualization interface.

@@

% \subsection{Pytracer workflow}
\pytracer adopts the three-step workflow shown in Figure~\ref{fig:workflow}. First, \pytracer uses monkey patching to replace the functions called by the application with a wrapper that saves their arguments and returned values in a trace file.  Second, it injects noise in the application, executes it multiple times, and aggregates the resulting trace files to compute summary statistics about each traced variable. Finally, a Plotly dashboard provides interactive visualizations highlighting unstable code sections.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/workflow.pdf}
    \caption{Pytracer workflow}
    \label{fig:workflow}
\end{figure}

@@

\pytracer dynamically instruments Python modules
without requiring source code modifications in the application.
To do so, it creates a new, instrumented version of each module that preserves its original attributes so that the application can transparently use it.
By default, \pytracer instruments all the attributes
in a given module except the special ones of the form \texttt{\_\_*\_\_}.
However, some attributes are not writable. In this case, \pytracer preservers the original attributes and warns the user. The user can also restrict the list of traced attributes through
an inclusion/exclusion mechanism. The main advantage of this instrumentation approach is the lack of required modification in the application source code, in contrast with decorator-based approaches. In addition, this technique is scalable and applicable in read-only environments such as Singularity containers or HPC clusters.

@@

\pytracer automatically instruments class instances since it instruments classes. However, it cannot instrument or instantiate classes that have read-only attributes --- in particular class \texttt{ufunc} ---, which is problematic given their pervasiveness in scientific computing. In particular, NumPy's universal functions include
vectorized elementary mathematical functions and other prevalent ones. Universal functions are implemented in C, which prevents their modification.

To overcome this issue, \pytracer wraps the class instance into a transparent wrap class (\texttt{twc}) that overloads
1) function \texttt{\_\_getattribute\_\_}, called when an instance attribute is accessed --- instrumented to return the attributes from the original class, and 2) function \texttt{\_\_call\_\_}, called when the function is called through the () operator --- instrumented with function \texttt{generic\_wrapper} . The \texttt{twc} returns the queried attribute of the instance instead of returning its attributes, allowing transparent access to the instance.  Finally, \pytracer overloads the builtin \texttt{type} function to preserve the type of the wrapped instance.

@@

Once traces are available, \pytracer sequentially parses them and computes the mean, standard deviation, and the number of significant digits (as in Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary statistics in an HDF5 file, a hierarchical format that facilitates visual browsing since only
required information are loaded when requested by the user.
This operation critically relies on the ordering of call input and output objects in the traces. Therefore, it assumes that the application is single-threaded and that the control flow does not depend on a random state. 
% Techniques to overcome this limitation are discussed in the discussion section.

@@

\subsection{Interactive visualization}
The third main component of \pytracer is its interactive trace visualizer built with Plotly Dash~\cite{plotly}, a Python framework for web visualization.
Plotly Dash abstracts the low-level details of web applications in a highly customizable development interface allowing to offload heavy computations to a server. 
Figure~\ref{fig:visu-layout} presents the general layout of the \pytracer visualizer, divided into five sub-components:
\begin{enumerate}
 \item Timeline view: This view displays the values of each input and output for a given function at a given invocation. By default, \pytracer shows the number of significant digits computed across the traces. The user can switch to the mean or the standard deviation. The visualizer uses an upper triangle for inputs and a lower triangle for outputs. \pytracer internally represents all values as NumPy arrays.    In the case of non-scalar values or types that cannot be trivially cast into NumPy arrays, \pytracer splits the value into several variables. Hence, \pytracer will transform a class containing $n$ floating-point parameters into $n$ variables. Each variable is prefixed with the original value and postfixed with the name of the parameter. 
    Finally, non-scalar values (vectors or matrices) are represented by their average and can be inspected with the Zoom view (see below).
 \item Gantt chart view: This view displays the call graph as a Gantt chart representing the instrumented functions. Each function call is represented with a bar.
    The time overlaps reflect the caller/callee relation.
    For example, in the figure, \texttt{function1} calls \texttt{function2} that itself calls \texttt{function4} and \texttt{function5}. 
    This view provides insights on the calling context for each function, which may influence numerical stability. 
 \item  Source code view: This view displays the source code file of the variable hovered in the timeline view. 
 \item Zoom view: This view provide details about non-scalar values. For example, a matrix may have instabilities around its diagonal but not in other regions.
    The visualizer also provides additional metrics such as array shape, infinite norm, or condition number.
\item  Functions list view: This view allows the user to select the functions plotted in the timeline view.
\end{enumerate}

@@

Table~\ref{tab:pytracer_results_summary} summarizes
\pytracer's outcome on SciPy and scikit-learn examples.
Columns \textit{trace} and \textit{parse} report on the tracing and trace aggregation steps with three possible outcomes: green for success, red for numerical error in the application, and orange for errors due to the fuzzy noise injection
that are not considered numerical errors.
% , like for instance index errors due to an index shift.
% \tristan{Explain what this is, in a setence here or with a subsubsection if it requires more}. 
Overall, the tracing and parsing of SciPy and scikit-learn examples in Random Rounding mode was entirely successful for 32/40 examples. The remaining examples failed due to numerical errors in the libraries, which we discuss hereafter. The instrumentation with Full MCA is more invasive and reduced the number of successful executions to 12/40.

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes the precision for the main outputs of SciPy and scikit-learn examples. 
We measured the precision in significant bits on the main example output by taking the element-wise mean for non-scalar values and, if the example produces many outputs, the average value among all of them. The numerical quality across all examples varies from 0 to 52, with an average of 44 significant bits. For both modes, we note that only two examples failed due to a numerical error 
% \tristan{can we see that in the table? if not, don't comment on it, it is puzzling. Could you show that in the table by using the same error symbols as in Table 2?} 
(\texttt{face\_recognition} for RR and \textit{Bayesian Ridge Regression} for Full MCA) even though we underestimate this number since Full MCA mode is more intrusive than RR mode and most of the examples produce error caused by the fuzzy noise injection.

\begin{table}[]
    \centering
    \small
    \begin{subfigure}[t]{\linewidth}
    \centering
    \begin{tabular}{|lll|c|c|c|c|c|c|}
    \hline
    \multicolumn{3}{|c}{ \multirow{2}{*}{Application} } & \multicolumn{3}{|c|}{Random Rounding (RR)} & \multicolumn{3}{c|}{Full MCA} \\
    \cline{4-9}
    & & & trace & parse & result & trace & parse & sigbits \\
    \hline
    %        &               RR                 &                 MCA               \\
    %        &  numpy         &   no numpy      &      numpy      &     no numpy    \\
    %        & run   & parse  &  run   &  parse &  run   & parse  & run    & parse  \\

>>>>

\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:} #1\color{black}\xspace}
\newcommand{\tristanmod}[2]{\color{orange}\sout{#1} #2\color{black}\xspace}
\newcommand{\gkmod}[2]{\color{purple}\sout{#1} #2\color{black}\xspace}

@@

\title{PyTracer: Automatically profiling numerical instabilities in Python}
\author{Yohan Chatelain$^1$, Nigel Young$^1$,  Gregory Kiar$^2$,  Tristan Glatard$^1$\\
$^1$Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada\\
$^2$Center for the Developing Brain, Child Mind Institute, New York, NY, USA}
    \date{}
\maketitle

\begin{abstract}
Numerical stability is a crucial requirement of reliable scientific computing. However, despite the pervasiveness of Python in data science, analyzing large Python programs remains challenging due to the lack of scalable numerical analysis tools available\gkmod{ for this language}{}. \gkmod{}{To fill this gap,}we developed \pytracer, a profiler to quantify numerical instability in Python applications. \pytracer \gkmod{automatically}{transparently?} instruments Python code to produce numerical traces and visualize them interactively in a Plotly dashboard. We designed \pytracer to \gkmod{work with any}{be agnostic to} numerical noise model \gkmod{including}{, allowing for tool evaluation through} Monte-Carlo Arithmetic, random rounding \gkmod{or}{,} random data perturbations\gkmod{}{, or structured noise for a particular application}. We illustrate \pytracer's capabilities by testing the numerical stability of key functions in \gkmod{the}{both} SciPy and Scikit-learn \gkmod{}{, two dominant} Python libraries \gkmod{}{for mathematical modeling}. Through these evaluations, we demonstrate \pytracer as a scalable, automatic, and generic framework for numerical \gkmod{evaluations}{profiling} in Python.
% Numerical stability is a crucial requirement of reliable scientific computing. However, despite the pervasiveness of Python in data science, analyzing large Python programs remains challenging due to the lack of scalable numerical analysis tools available for this language. To address this issue, we developed \pytracer, a profiler to quantify numerical instability in Python applications. \pytracer automatically instruments Python code to produce numerical traces and visualize them interactively in a Plotly dashboard.
% \pytracer is designed to work with any type of numerical noise model, including Monte-Carlo Arithmetic, random rounding or random data perturbations. We illustrate \pytracer's capabilities by testing the numerical stability of key functions in the SciPy and Scikit-learn Python libraries. Finally, \pytracer slows down executions by only $3.5 \times$ on average.
% \tristan{Add one sentence to summarize the findings} 
\end{abstract}

@@

The scientific Python ecosystem \gkmod{is}{has become} a central component of data
analyses \gkmod{}{in recent years,} owing to its rich offering of data manipulation, array programming,
numerical analysis, and visualization tools. In particular, libraries such
as NumPy~\cite{harris2020array}, SciPy~\cite{virtanen2020scipy}, scikit-learn~\cite{pedregosa2011scikit} or PyTorch~\cite{paszke2019pytorch} are used in hundreds of publications \gkmod{every}{each} year, \gkmod{providing}{and serve as} a reference set of high-quality open-source core scientific libraries. Researchers have built numerous domain-specific software tools from this ecosystem, leveraging Python's simplicity and \gkmod{expressivity}{flexibility}.

Numerical stability is a crucial requirement of reliable scientific data
processing. In unstable analyses, small numerical perturbations introduced by data noise, software and hardware updates, or parallelization \gkmod{lead to}{have potential to introduce} substantial deviations in final results, \gkmod{and}{leading to} potentially different scientific conclusions, \gkmod{ultimately} threatening the reliability of computational analyses. To \gkmod{address}{prevent} this \gkmod{issue}{}, numerical stability analyses need to be conducted systematically, however, no practical tool currently exists to conduct such analyses for Python programs.

We present PyTracer, a numerical stability \gkmod{analyzer}{profiling and visualization tool} for Python programs.
\pytracer adopts a dynamic approach that evaluates numerical stability using program execution traces, and is applicable to large programs without manual intervention. \gkmod{}{The transparent nature of \pytracer is particularly valuable} in contrast \gkmod{}{to} theoretical approaches based on condition numbers or backward error analysis\gkmod{}{, which} require detailed modeling of the program and its numerical implementation. Similarly, static code analysis techniques hardly scale to large codebases, particularly in dynamically-typed languages such as Python.

\pytracer estimates the variability of floating-point variables by combining program traces obtained with different numerical perturbations, which requires (1) generating numerically-perturbed executions, (2) tracing floating-point computations along program executions, and (3) visualizing stability evaluations. \pytracer addresses these requirements with (1) a ``fuzzy" Python interpreter instrumented with Monte-Carlo arithmetic, (2) a dynamic instrumentation of Python functions, modules, and classes through meta-programming, (3) an interactive Plotly dashboard to visualize stability measures throughout program executions.

@@

Several tools have been developed to assess numerical quality, \gkmod{}{and can be} divided into static and dynamic approaches. Static approaches typically analyze source code, while dynamic approaches trace floating-point computations throughout executions. A detailed review of these approaches is presented in~\cite{cherubin2020tools}. 
\pytracer adopts the dynamic approach due to its scalability to large \gkmod{existing}{} code bases. Dynamic approaches typically trace floating-point computations, detect instabilities in traces, and visualize summary statistics about instabilities. This section reviews existing approaches for each of these steps.

@@

\label{sec:soa}
\subsection{Numerical tracing techniques}
Numerous tools exists to trace floating-point computations for C, C++, or FORTRAN programs due to the prevalence of these languages in High-Performance Computing (HPC). 
The main tracing techniques are \textit{source-to-source}, \textit{compiler-based transformations}, and \textit{dynamic binary instrumentation}.

\gkmod{}{Note: this paragraph just reads like a list with no real interpretation/commentary. I think it would be better to state the strengths of this technique, cite examples, and then state limitations/why it isn't the only technique used.}
The source-to-source technique requires a rewriting of the application to modify floating-point types. 
CADNA~\cite{jezequel2008cadna} is a library for C, C++, and Fortran implementing \gkmod{the}{} CESTAC~\cite{vignes1993stochastic} stochastic arithmetic. Shaman~\cite{demeure_phd} is a C++ library that uses a first-order error model to propagate numerical error. 
MCAlib~\cite{frechtling2015mcalib} is a C library that implements the Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} using the MPFR~\cite{fousse2007mpfr} library.
Finally, the work in~\cite{tang2016software} proposed a source-to-source framework for executing targeted code in infinite and fixed precision with and without  stochastic arithmetic.

\gkmod{}{Note: similar to the above, this is just a list rather than a highlight of how the compiler-based approach can be more scalable than source-to-source because it requires less manual intervention into a codebase (e.g.), then have your list, and finally say why it is still limited/not a global fix to this problem.}
The compiler-based approach instruments floating-point expressions at compile time. 
Verificarlo~\cite{verificarlo} is a compiler that supports different types of arithmetic instrumentations, including MCA. 
The work in~\cite{bao2013fly} modified the GNU Compiler Collection (GCC) to track local floating-point errors across executions. pLiner~\cite{guo2020pliner} is a root-cause analyzer based on the Clang compiler that detects floating-point operations responsible for result variability using a source-to-source transformation at the Abstract Syntax Tree (AST) level to rewrite parts of code with higher precision. 
PFPSanitizer~\cite{chowdhary2020debugging,chowdhary2021parallel} is a compiler that uses parallel shadow execution to detect numerical issues using higher precision.
FLiT~\cite{sawaya2017flit} is a framework to detect variability induced by compilers and their optimizations.
The work in~\cite{wang2012development} proposes a numerical debugger based on GDB~\cite{stallman1988debugging} for Discrete Stochastic Arithmetic (DSA) on FPGA as an Eclipse plugging. Similarly, Cadtrace~\cite{jezequel2008cadna} and Shaman propose a GDB-based tool to use the CADNA and Shaman libraries with GDB, respectively.

@@

\gkmod{}{Note: I like this comparison here, but do think that the comments above about providing some more info when each is mentioned would still help this section.}
Being agnostic to the programming language is a serious advantage of the DBI method compared to source-to-source and compiler-bases methods. However, working at the binary level makes it difficult to access high-level information needed to debug and understand the source-code logic. Conversely, the source-to-source approach provides a fine-grained control on the analyzed code, but it lacks scalability, as rewriting large codes can be a tedious task. Finally, the compiler-based approach takes advantage of the best of both by being automatic and having access to high-level information. However, like source-to-source, the compiler-based method is not suitable for analyzing closed-source libraries.

@@

Three main approaches exist to detect numerical instabilities: stochastic arithmetic, uncertainty or sensitivity analysis~\cite{helton2006survey}, and random seed analysis~\cite{click2011quality}.
%\tristan{add key references for uq/sa if any}
\pytracer focuses on stochastic arithmetic since it does not make any assumption about 
the noise model, in contrast to sensitivity analysis. \gkmod{}{Q: isn't \pytracer agnostic to noise model/method for instability detection? Can't we just re-run the tool multiple times under whatever model for detecting instabilities that we choose and it will work all the same? If so, then I may reword this to say that pytracer works for all, and for our usecases here we use SA because X, Y, and Z...}
%\tristan{due to...}
Stochastic arithmetic leverages randomness to estimate numerical instabilities coming from the use of floating-point representations. The main idea is to treat round-off errors 
% \tristan{only round-off errors or all numerical errors?} 
as random variables and characterize them statistically.

@@

Stochastic arithmetic quantifies numerical error using the number of significant digits $s$ estimated among the sampled values. A common formula to determine this number from MCA samples is \gkmod{the one reported}{presented} in~\cite{parker1997monte}:
\begin{equation}
s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a variable sampled with MCA.  
Sohier et al.~\cite{sohier2018confidence} recently provided a generalization of this formula to include confidence intervals.

@@

Visualization \gkmod{}{Note: You kind of jump right in to this section. Visualization of what in particular? Why do we want (need?) to visualize this sort of data?} covers a wide range of numerical simulation domains, from general-purpose numerical visualization with tools
such as VTK~\cite{schroeder2000visualizing} to topological data analysis~\cite{tierny2018topological}, including matrix~\cite{wu2008matrix} or tensor~\cite{kindlmann2006diffusion} visualization.
Despite the large number of existing tools to identify numerically unstable sections of source code, few allow the user to have a global view of the numerical precision of their computations or to compare several executions of the same program under various perturbations.

Veritracer~\cite{chatelain2018veritracer} was the first tool to propose an embedded visualizer of numerical stability over the execution of C, C++, and Fortran code. Veritracer extends Verificarlo~\cite{verificarlo} with contextual information about traced variables, which allows the user to identify instabilities quickly.
However, Veritracer neither offers a dynamic interface nor includes information about high-level structures such as \texttt{struct} or \texttt{array}.
FPAvizual~\cite{gu2014fpavisual} visualizes the effects of floating-point arithmetic but only for educational purposes \gkmod{}{Q: What does this mean?}.
Anteater~\cite{faust2019anteater} proposes a visualizer for Python code, but as mentioned previously, it only deals with native Python types, and the user needs to select variables to trace manually. Anteater provides an insightful reflection 
around visual debugging beyond floating-point analysis, which has inspired the design of \pytracer's visualization interface.

@@

% \subsection{Pytracer workflow}
\pytracer adopts the three-step workflow shown in Figure~\ref{fig:workflow}. First, \pytracer uses \gkmod{monkey patching}{\textit{monkey patching}} to replace the functions called by the application with a wrapper that saves their arguments and returned values in a trace file \gkmod{}{Q: This is conceptually similar to a decorator, no? If so, perhaps it is worth mentioning}. Second, it injects noise in the application \gkmod{}{Q: \pytracer doesn't actually inject noise, does it? You can just run your tool however you want multiple times}, executes it multiple times, and aggregates the resulting trace files to compute summary statistics about each traced variable. Finally, a Plotly dashboard provides interactive visualizations highlighting unstable code sections.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/workflow.pdf}
    \caption{Pytracer workflow \gkmod{}{Note: replace "Execution" with "Multiple Independent Executions", and maybe move it to be between the gears and the pkl files.}}
    \label{fig:workflow}
\end{figure}

@@

\pytracer dynamically instruments Python modules
without requiring source code modifications in the application.
To do so, it creates a new, instrumented version of each module that preserves its original attributes so that the application can transparently use it. \gkmod{}{Note: this form of monkey patching is basically global and transparent decorating, right?}
By default, \pytracer instruments all the attributes
in a given module except the special ones of the form \texttt{\_\_*\_\_}\gkmod{}{, which is common for reserved Python methods}.
However, some attributes are not writable. In this case, \pytracer preservers the original attributes and warns the user. The user can also restrict the list of traced attributes through
an inclusion/exclusion mechanism. The main advantage of this instrumentation approach is the lack of required modification in the application source code, in contrast with decorator-based approaches. In addition, this technique is scalable and applicable in read-only environments such as Singularity containers or HPC clusters.

@@

\pytracer automatically instruments class instances since it instruments classes. However, it cannot instrument or instantiate classes that have read-only attributes --- in particular class \texttt{ufunc}, --- which is problematic given their pervasiveness in scientific computing. In particular, NumPy's universal functions include
vectorized elementary mathematical functions and other prevalent \gkmod{ones}{Q: prevalent whats?}. Universal functions are implemented in C, which prevents their modification.

To overcome this issue, \pytracer wraps the class instance into a transparent wrap class (\texttt{twc}) that overloads
1) function \texttt{\_\_getattribute\_\_}, called when an instance attribute is accessed --- instrumented to return the attributes from the original class, and 2) function \texttt{\_\_call\_\_}, called when the function is called through the () operator --- instrumented with function \texttt{generic\_wrapper}. \gkmod{}{Note: the previous sentence is a bit hard to follow; the word "called" is used a lot so I got a bit lost.} The \texttt{twc} returns the queried attribute of the instance instead of returning its attributes, allowing transparent access to the instance.  Finally, \pytracer overloads the builtin \texttt{type} function to preserve the type of the wrapped instance.

@@

Once traces are available, \pytracer sequentially parses them and computes the mean, standard deviation, and the number of significant digits (as in Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary statistics in an HDF5 file, a hierarchical format that facilitates visual browsing since only
required information are loaded when requested by the user. \gkmod{}{Q: Is there a reason we don't also use HDF for the raw traces? Lack of native compression?}
This operation critically relies on the ordering of call input and output objects in the traces. Therefore, it assumes that the application is single-threaded and that the control flow does not depend on a random state. 
% Techniques to overcome this limitation are discussed in the discussion section.

@@

\subsection{Interactive visualization}
The third main component of \pytracer is its interactive trace visualizer built with Plotly Dash~\cite{plotly}, a Python framework for web visualization.
Plotly Dash abstracts the low-level details of web applications in a highly customizable development interface allowing to offload heavy computations to a server \gkmod{}{Q: does it support using a remote server with forwarding, in the event that your local system isn't that beefy?}. 
Figure~\ref{fig:visu-layout} presents the general layout of the \pytracer visualizer, divided into five sub-components:
\begin{enumerate}
 \item Timeline view: This view displays the values of each input and output for a given function at a given invocation. By default, \pytracer shows the number of significant digits computed across the traces. The user can switch to the mean or the standard deviation \gkmod{}{Q: mean/stdev of what? the inputs/outputs? But what if they are on a totally different scale (e.g. log(0.0001)}. The visualizer uses an upper triangle for inputs and a lower triangle for outputs. \pytracer internally represents all values as NumPy arrays. In the case of non-scalar values or types that cannot be trivially cast into NumPy arrays, \pytracer splits the value into several variables. Hence, \pytracer will transform a class containing $n$ floating-point parameters into $n$ variables. Each variable is prefixed with the original value and postfixed with the name of the parameter. \gkmod{}{Q: how is this managed for the sake of interpretability? E.g. if one of the components has very low significance and another few are fine, how can I easily identify where in my sequence we have the drop in precision?}
    Finally, non-scalar values (vectors or matrices) are represented by their average and can be inspected with the Zoom view (see below).
 \item Gantt chart view: This view displays the call graph as a Gantt chart representing the instrumented functions. Each function call is represented with a bar.
    The time overlaps reflect the caller/callee relation.
    For example, in the figure, \texttt{function1} calls \texttt{function2} that itself calls \texttt{function4} and \texttt{function5}. 
    This view provides insights on the calling context for each function, which may influence numerical stability. \gkmod{}{Note: this is a non-standard layout for gantt charts... normally the width/overlap is determined by time of start + duration. Is that information contained somewhere in here? If not, will a very fast function that calls sub-functions, say, 1000 times look longer than a very slow function that doesn't call any subfunctions?}
 \item  Source code view: This view displays \gkmod{}{a local snapshot of} the source code file of the variable hovered in the timeline view. 
 \item Zoom view: This view provide details about non-scalar values. For example, a matrix may have instabilities around its diagonal but not in other regions.
    The visualizer also provides additional metrics such as array shape, infinite norm, or condition number.
\item  Functions list view: This view allows the user to select the functions plotted in the timeline view.
\end{enumerate}

@@

Table~\ref{tab:pytracer_results_summary} summarizes
\pytracer's outcome on SciPy and scikit-learn examples.
Columns \textit{trace} and \textit{parse} report on the tracing and aggregation steps with three possible outcomes: green for success, red for numerical error in the application, and orange for errors due to the fuzzy noise injection
that are not considered numerical errors. \gkmod{}{Q: What does "success" or the other labels actually mean? What is the threshold for numerical errors, e.g.}
% , like for instance index errors due to an index shift.
% \tristan{Explain what this is, in a setence here or with a subsubsection if it requires more}. 
Overall, the tracing and parsing of SciPy and scikit-learn examples in Random Rounding mode was entirely successful \gkmod{}{, indicating on average XX loss in numerical precision,} for 32/40 examples. The remaining examples failed due to numerical errors in the libraries, which we discuss hereafter. The instrumentation with Full MCA is more invasive and reduced the number of successful \gkmod{}{Q: Does this mean the 28 failed, or completed but poorly? In general, the way the states are being referred to so far in the results are rather ambiguous} executions to 12/40.

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes the precision for the main outputs of SciPy and scikit-learn examples. 
We measured the precision in significant bits on the \gkmod{main}{final primary} example output by taking the element-wise mean for non-scalar values and, if the example produces many outputs, the average value among all of them. The numerical quality across all examples varies from 0 to 52, with an average of 44 significant bits. For both modes, we note that only two examples failed due to a numerical error 
% \tristan{can we see that in the table? if not, don't comment on it, it is puzzling. Could you show that in the table by using the same error symbols as in Table 2?} 
(\texttt{face\_recognition} for RR and \texttt{Bayesian Ridge Regression} for Full MCA) even though this number is underestimated since Full MCA mode is more intrusive than RR mode and most of the examples produce error caused by the fuzzy noise injection. \gkmod{}{Note: I don't understand what is being said in the rest of the sentence following the parentheses}

\begin{table}[]
    \centering
    \small
    \begin{subfigure}[t]{\linewidth}
    \centering
    \begin{tabular}{|lll|c|c|c|c|c|c|}
    \hline
    \multicolumn{3}{|c}{ \multirow{2}{*}{Application} } & \multicolumn{3}{|c|}{Random Rounding (RR)} & \multicolumn{3}{c|}{Full MCA} \\
    \cline{4-9}
    & & & trace & parse & sigbits & trace & parse & sigbits \\
    \hline
    %        &               RR                 &                 MCA               \\
    %        &  numpy         &   no numpy      &      numpy      &     no numpy    \\
    %        & run   & parse  &  run   &  parse &  run   & parse  & run    & parse  \\
